Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
Soundness edits

 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked, since by assumption such a program
 exists.  Since the first-order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.
\end{proof}

@@

\begin{corollary}
 Every finite subset $A \subseteq B$ is computable by a finite $\mathcal{L}$-program
 by setting $X = B, Y = 2$ in Lemma~\ref{lem:l-func}.
\end{corollary}

@@

\begin{proof}
%
Let $v_1, \ldots, v_k$ be the variables of the program $P$ under analysis,
and let each be $b$ bits wide.  Its state space $\mathcal{S}$ is then of
size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
exists iff $P$ terminates.  Without loss of generality,
$\mathcal{D}$ is a well-founded total order.  Since $R$ is injective, we have that $\|
\mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
\mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to
\mathcal{D'}$ with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting
$R' = R|_\mathcal{S}$, i.e.~$R'$ is just the restriction of $R$ to
$\mathcal{S}$.  Since $\mathcal{S}$ already comes equipped with a natural
well ordering we can also construct $R'' = \iota \circ R'$ where $\iota:
\mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from
$\mathcal{D'}$ to $\mathcal{S}$.  So assuming that $P$ terminates, there is
some ranking function $R''$ that is just a permutation of $\mathcal{S}$.  If
the number of variables $k > 1$ then in general the ranking function will be
lexicographic with dimension $\leq k$ and each co-ordinate of the output
being a single $b$-bit value.

@@

\begin{proof}
 A non-termination proof is a triple $\langle N, C, x_0 \rangle$ where $N \subseteq \mathcal{S}$ is
 a (finite) recurrence set and $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
 a successor for each $x \in N$.  Since $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
 $N$ and $C$ are computed by finite $\mathcal{L}$-programs.
\end{proof}

@@

 So in either case ($L$ terminates or does not), there is a witness in $\mathcal{L}$ satisfying
 formula {\bf [GT]}: $\phi \vee \psi$.
\end{proof}

>>>>

 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 on whatever set of inputs are currently being tracked, since by assumption such a program
 exists.  Since the first-order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.
\end{proof}

@@

\begin{corollary}
 Every finite subset $A \subseteq B$ is computable by a finite $\mathcal{L}$-program
 by setting $X = B, Y = 2$ in Lemma~\ref{lem:l-func} and taking the
 resulting function to be the characteristic function of $A$.
\end{corollary}

@@

\begin{proof}
%
Let $v_1, \ldots, v_k$ be the variables of the program $P$ under analysis,
and let each be $b$ bits wide.  Its state space $\mathcal{S}$ is then of
size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
exists iff $P$ terminates.  Without loss of generality,
$\mathcal{D}$ is a well-founded total order.  Since $R$ is injective, we have that $\|
\mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
\mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to
\mathcal{D'}$ with $ \| \mathcal{D'} \| = \| \mathcal{S} \|$ by just setting
$R' = R|_\mathcal{S}$, i.e.~$R'$ is just the restriction of $R$ to
$\mathcal{S}$.  Since $\mathcal{S}$ already comes equipped with a natural
well ordering we can also construct $R'' = \iota \circ R'$ where $\iota:
\mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from
$\mathcal{D'}$ to $\mathcal{S}$.  So assuming that $P$ terminates, there is
some ranking function $R''$ that is just a permutation of $\mathcal{S}$.  If
the number of variables $k > 1$ then in general the ranking function will be
lexicographic with dimension $\leq k$ and each co-ordinate of the output
being a single $b$-bit value.

@@

\begin{proof}
 A non-termination proof is a triple $\langle N, C, x_0 \rangle$ where
 \mbox{$N \subseteq \mathcal{S}$} is a (finite) recurrence set and
 $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
 a successor for each $x \in N$.  $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
 $N$ and $C$ are computed by finite $\mathcal{L}$-programs and $x_0$ is just a ground term.
\end{proof}

@@

 So in either case ($L$ terminates or does not), there is a witness in $\mathcal{L}$ satisfying
 $\phi \vee \psi$, which is an instance of {\bf [GT]}.
\end{proof}
